Modelling the travel time of transit vehicles in \rt through a GTFS-based road network using GPS vehicle locations

### 2018-11-01 anzjs2018/anzjs2018.tex
Clean up dir

\begin{document}
\cleanlookdateon
\runningheads{Modeling transit vehicles in real-time}{TOM~ELLIOTT AND THOMAS~LUMLEY}
\title{A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network}
\author{Tom Elliott\corrauth~and~Thomas~Lumley}
\affiliation{University of Auckland}
\address{
    Department of Statistics, University of Auckland, 
    Auckland 1142, New Zealand\\
    Email: \texttt{tom.elliott@auckland.ac.nz}
}
\ack{
    We would like to thank Auckland Transport for providing free,
    open access to the \rt data; and the Center for e-Research at the
    University of Auckland for providing and maintaining the virtual
    machine used in our work.
}

\begin{abstract}
Predicting the arrival time of a transit vehicle involves not only
knowledge of its current position and schedule adherence,
but also of traffic conditions along the remainder of the route.
Road networks are dynamic, 
and can change from free-flowing to highly congested very quickly,
which impacts the arrival time of transit vehicles,
particularly buses which often share the road with other vehicles,
so reliable predictions need to account for \rt and future traffic conditions.
The first step in this process is to construct a framework with which road state
(traffic conditions) can be estimated using \rt transit vehicle position data.
We propose a framework which implements a vehicle model using a particle filter
to estimate travel times along roads,
used by a second model to estimate \rt traffic conditions which can
later be used for arrival time prediction.
We demonstrate the \rt feasibility of such an approach in \rt,
and detail our implementation in R and C++.
Development and testing took place in Auckland, New Zealand,
with future work planned to incorporate historical data into the network model
to improve the reliability of ETAs.
\end{abstract}

>>>>

\begin{document}
\cleanlookdateon
\runningheads{Modelling the travel time of transit vehicles}{TOM~ELLIOTT AND THOMAS~LUMLEY}
\title{
    Modelling the travel time of transit vehicles in \rt through a GTFS-based road network
    using GPS vehicle locations
}

% Modelling transit vehicles to estimate the \rt traffic conditions in a public transport network}
\author{Tom Elliott\corrauth~and~Thomas~Lumley}
\affiliation{University of Auckland}
\address{
    Department of Statistics, University of Auckland, 
    Auckland 1142, New Zealand\\
    Email: \texttt{tom.elliott@auckland.ac.nz}
}
\ack{
    We would like to thank Auckland Transport for providing free,
    open access to the \rt data; and the Center for e-Research at the
    University of Auckland for providing and maintaining the virtual
    machine used in our work.
}

\begin{abstract}
Predicting the arrival time of a transit vehicle involves not only
knowledge of its current position and schedule adherence,
but also of traffic conditions along the remainder of the route.
Road networks are dynamic, 
and can change from free-flowing to highly congested very quickly,
which impacts the arrival time of transit vehicles,
particularly buses which often share the road with other vehicles,
so reliable predictions need to account for \rt and future traffic conditions.
The first step in this process is to construct a framework with which road state
(traffic conditions) can be estimated using \rt transit vehicle position data.
Our proposed framework implements a vehicle model using a particle filter
to estimate road travel times,
which are used in a second model to estimate \rt traffic conditions.
Development and testing took place in Auckland, New Zealand,
although attempts were made at generalising each component so 
it will be compatible with other public transport systems around the world.
We demonstrate the \rt feasibility and performance of our approach in real-time,
where a combination of R and C++ was used to obtain the necessary performance results.
Future work will use these estimated traffic conditions 
in combination with historical data to obtain reliable
arrival time predictions of transit vehicles.
\end{abstract}

### 2018-11-01 anzjs2018/sections/01_introduction.tex
Clean up dir

In public transport, \rt information (RTI), such as estimated times of arrival (ETAs),
keeps transport users informed about their journey,
and potentially make decisions on the best route to take.
It has also been shown that perceived waiting time is less
when arrival time information is available \citep{TCRP_2003b}.
Unfortunately, here in Auckland, RTI is highly unreliable,
and succumbs to several problems that public transport users
are all too familiar with.
One of these is inaccurate ETAs,
which tend to either increase or decrease over time as the bus gets progressively later or earlier
due to traffic congestion, or  because the schedule is poorly calibrated.
Another is that buses are shown as \emph{on-time} if they are not present
in the \rt system,
regardless of whether this is due to technical problems,
or the bus failing to show up;
in either case, schedule times are used to determine ETAs and, 
if the bus doesn't show up on time,
the service is removed from the \rt board 
and we get what passengers have described as ``Ghost Buses''.

Arrival time prediction is only as reliable as the underlying model,
and while a lot of work has gone into developing public transport vehicle models
\citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
in many public transport systems no formal model is used.
Instead, ETAs are solely based on the scheduled arrival time
adjusted by the vehicle's delay at the most recent stop, if available.
This assumes that the schedule is valid,
and that there is no unusual congestion along the route,
neither of which are valid assumptions,
particularly in our test area of Auckland, New Zealand,
where infrastructure for buses (such as priority lanes) is limited.
This behaviour is exaggerated when there is a large distance between stops,
as the ETA is only updated at stops so there can be a long time between updates.

To avoid the problems discussed above,
a more robust modelling and prediction framework is required.
This should consist of a robust vehicle model to estimate the position and speed
of transit vehicles from \rt GPS data;
and a way to incorporate road congestion information 
into the arrival time predictions.
Several vehicle modelling approaches were explored, 
including the Kalman filter \citep{Dailey_2001,Cathey_2003},
machine learning models \citep{Yu_2006,Chang_2010},
and a particle filter \citep{Hans_2015}.
The last of these was chosen for several reasons
which are discussed in Section~\ref{sec:pf}.

The next step is to construct a method of estimating traffic conditions.
This can be done by using the travel time of preceding buses along the road
to predict travel times for the current bus.
\cite{Yu_2010} obtained improved prediction accuracy by using travel times
from preceding buses servicing the same route.
\cite{Hans_2015} used a similar approach,
instead referring to headway, the time between consecutive vehicles
at a point on the route.
However, this approach only works well for high frequency routes.
\cite{Yu_2011} showed further improvements by combining travel times 
from several routes which converged,
but is limited to predefined, converging routes.
In general, however, no comprehensive network modelling approach has been proposed using
solely GPS position data to accounting for congestion when estimating arrival times.
We describe a simple method of constructing a road network
from GTFS data (section~\ref{sec:gtfs}),
which allows all transit vehicles travelling along a road to contribute to its state,
regardless of which route they are serving.

In this paper, we describe a \rt vehicle and network state
model which uses the vehicles travelling throughout the transit road network
to update congestion information in real-time.
As the vehicle model is implemented using a particle filter,
a computationally intensive method,
we demonstrate its feasibility as a valid real-time option.
The vehicle model provides estimates of travel times along roads,
which can be used to update the road network's state.
Future work is planned to use this travel time information 
in combination with historical data to develop an arrival time prediction model.

>>>>

In public transport, \rt information (RTI), such as estimated times of arrival (ETAs),
keeps transport users informed about their journey,
and enabled them to make decisions about the best route.
It has also been shown that perceived waiting time is less
when arrival time information is available \citep{TCRP_2003b}.
Unfortunately, here in Auckland, RTI is highly unreliable,
and succumbs to several problems that public transport users
are all too familiar with.
Inaccurate ETAs are a major source of frustration,
as they tend to either increase or decrease over time 
as the bus gets progressively later or earlier due to traffic congestion, 
or simply due to poor schedule calibration.
Furthermore, buses are shown as \emph{on-time} 
if they are not present in the \rt system and have not been cancelled,
so ETAs are based solely on scheduled arrival times;
if the bus does not show up on time
the service is removed from the \rt board,
leaving passengers unsure as to when---if at all---their bus will show up,
a phenomenon termed by bus users as ``ghost buses''.

Arrival time prediction is only as reliable as the underlying model,
and while a lot of work has gone into developing public transport vehicle models
\citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
in many public transport systems no formal model is used.
Instead, ETAs are solely based on the scheduled arrival time
adjusted by the vehicle's delay at the most recent stop, if available.
This assumes that the schedule is valid,
and that there is no unusual congestion along the route,
neither of which are valid assumptions,
particularly in our test area of Auckland, New Zealand,
where infrastructure for buses (such as priority lanes) is limited.

To avoid the problems mentioned above,
a more robust modelling and prediction framework is required.
This should consist of a robust vehicle model to estimate the position and speed
of transit vehicles from \rt GPS data,
and a way to incorporate road congestion information 
into the arrival time predictions.
Several vehicle modelling approaches were explored, 
including the Kalman filter \citep{Dailey_2001,Cathey_2003},
machine learning models \citep{Yu_2006,Chang_2010},
and a particle filter \citep{Hans_2015}.
We chose to use a particle filter for several reasons,
which are discussed in Section~\ref{sec:pf}.

The next step involves estimating traffic conditions throughtout the transport network
using the information obtained from the GPS data. 
\cite{Yu_2010} improved prediction accuracy by using the travel times
of buses travelling along the same route.
A similar method presented by \cite{Hans_2015}
used headway, the time between consecutive vehicles at a point on the route,
as a predictor of travel time.
Since these approaches only work well on high frequency routes,
\cite{Yu_2011} showed further improvements by combining travel times 
from several routes;
however, this was limited to predefined converging routes.
In general, however, no comprehensive network modelling approach has been proposed using
solely GPS position data to model and account for congestion when estimating arrival times.

In Section~\ref{sec:models} of this paper, 
we describe the \rt vehicle and network state models
used to estimate traffic congestion using the transit vehicles
travelling through the network in real-time,
and in Section~\ref{sec:rt} we demonstrate the feasibility of using a particle filter
in \rt to model transit vehicles,
and evaluate its performance.
First, however, we descibe how we construct a
road network from GTFS data in Section~\ref{sec:gtfs}.

### 2018-11-01 anzjs2018/sections/02_gtfs.tex
Clean up dir

There are two components to GTFS.
The first, \emph{GTFS static}, includes information about
\begin{itemize}
\item \emph{stops}, a physical location where passengers can embark and disembark the vehicle;
\item \emph{routes}, a sequence of two or more stops displayed as a single service;
\item \emph{trips}, an instance of a route occurring at a specific time of day;
\item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops; 
\item \emph{shapes}, the sequence of points defining a vehicle's path along a route
\end{itemize}
\citep{GoogleDevelopers_2006}.
The second component is \emph{GTFS realtime},
which is only available in a subset of the providers due to the requirement of 
on board GPS tracking devices and a central server.
It provides a standardised format for sharing vehicle positions and trip delays,
and are typically accessed by developers via an API can be used in \rt applications.

As mentioned in Section~\ref{sec:intro},
there are some major issues with the current prediction method in Auckland.
These are almost directly attributed to GTFS-realtime trip updates,
which are currently the sole source of data for ETAs.
Trip updates are reported whenever a vehicle arrives or departs a stop,
and includes the delay between the scheduled and actual arrival times,
and is propagated to all future stops to adjust their ETAs.
This assumes that the schedule is well calibrated and the time between scheduled arrivals
is representative of the real-world travel time between stops. 
Therefore, an alternative estimation approach using \rt travel times 
between stops is more adequate.

@@

The primary predictor of arrival time is 
the travel time along roads between where the bus is \emph{now},
and the stop at which the passenger is waiting.
In most applications, however, this vital information is unavailable,
at least directly,
so here we construct a transit road network consisting of intersections
and the roads between them.
By generalising to physical roads,
we remove any route-dependency,
which increases the quantity of data available for each road.
Each route is then represented as a sequence of road segments,
each with an associated travel time.

The transit network is constructed by splitting each route
into spatially identifiable segments,
each representative of a physical road.
The simplest way to do this is to use bus stops as nodes in the network
(Figure~\ref{fig:network_creation_1}),
and connecting roads as edges.
In this way, routes that service the same subsequence of stops
will all contribute to congestion information for the connecting roads,
as shown in Figure~\ref{fig:network_creation_2}.
There are several drawbacks to this method,
for example routes merge at physical intersections, not bus stops,
so some overlap will occur as shown.
However, this is minimal, and our simplification is sufficient for the current work.

\begin{figure}[tb]
    \centering
    \begin{subfigure}{0.7\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_1.pdf}
        \caption{Raw GTFS route shapes}
        \label{fig:network_creation_1}
    \end{subfigure} \\
    \begin{subfigure}{0.7\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_2.pdf}
        \caption{GTFS-based road network}
        \label{fig:network_creation_2}
    \end{subfigure}
    \caption{
        Constructing a route network from the stops (dots) of three separate
        routes (lines), as shown in (a).
        Routes using the same subsequence of stops are assigned the same
        road segments, so the road network shown 
        consists of nodes (stops) and edges (roads),
        as shown in (b), where the thickness of each lines
        corresponds to how many routes share that road.
    }
    % Construction of a route network involves combining routes at nodes %
    %     (here we are using stops) which are connected by edges (roads). %
    %     In (a), the three unique routes, represented by different line types, clearly %
    %     overlap in several places. In (b) these have been merged, and the width of each line %
    %     represents how many routes use that link.}
    \label{fig:network_creation}
\end{figure}

@@

GTFS-realtime provides the current positions of vehicles in a transit network.
The data consists of the time $t_k$ that the observation was made,
the GPS position of the vehicle, $\bY_k$, 
and information about the trip being serviced.
In Auckland, vehicle positions are updated with a frequency of anywhere between 10~seconds and several minutes,
so there is often a lot of uncertainty about a vehicle's intermediate trajectory,
particularly when there are one or more bus stops along the way.
It is also possible for a bus to remain stationary,
for example due to heavy congestion,
so the number of possible trajectories rapidly increases with 
the time between observations.

One important consideration regarding Auckland Transport's \rt implementation is that
buses are programmed to report their location when arriving at or departing from
bus stops, as well as some major intersections.
To further complicate this,
these positions can be preemptive,
(i.e., the bus is almost there, but not quite),
so consecutive observations can show a bus travelling backwards.
To handle this, we compute the approximate distance travelled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the current state is rejected and the vehicle reverted
to its previous state before continuing.
This way, we avoid degeneration of the particle filter (Section~\ref{sec:model_perf})
at the cost of losing one observation that is most likely invalid anyway.

>>>>

There are two components to GTFS.
The first, \emph{GTFS static}, includes information about
\begin{itemize}
\item \emph{stops}, a physical location where passengers can embark and disembark the vehicle;
\item \emph{routes}, a sequence of two or more stops displayed as a single service;
\item \emph{trips}, an instance of a route occurring at a specific time of day;
\item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops; 
\item \emph{shapes}, the sequence of points defining a vehicle's path along a route
\end{itemize}
\citep{GoogleDevelopers_2006}.
The second component is \emph{GTFS-realtime},
which is only available in a subset of the providers due to the requirement of 
on board GPS tracking devices and a central server.
It provides a standardised format for sharing vehicle positions and trip delays,
and are typically accessed by developers via an API can be used in \rt applications.

As mentioned in Section~\ref{sec:intro},
there are some major issues with the current prediction method in Auckland.
These are almost directly attributed to \emph{GTFS-realtime} trip updates,
which are currently the sole source of data for ETAs.
Trip updates are reported whenever a vehicle arrives or departs a stop,
and includes the delay between the scheduled and actual arrival times,
which is propagated to all future stops to update their ETAs.
This assumes that the schedule is well calibrated 
and the scheduled travel time between stops
is representative of the real-world travel time between them. 
Therefore, an alternative estimation approach that uses \rt travel times 
between stops would seem more adequate.

@@

The primary predictor of arrival time is 
the travel time along roads between where the bus is \emph{now},
and the stop at which the passenger is waiting.
In most applications, however, this vital information is unavailable,
at least directly,
so here we describe how we constructed a transit road network consisting of intersections
and the roads between them.
Generalising to physical roads removes route dependencies,
which increases the quantity of data available for each road
and allows each route to be represented as a sequence of road segments,
which have individual travel time distributions (Section~\ref{sec:kf}).

The transit network was constructed by splitting each route
into spatially identifiable segments,
each representative of a physical road,
by using bus stops as nodes in the network
and the connecting roads as edges
(Figure~\ref{fig:network_creation}).
In this way, routes that service the same subsequence of stops
all contribute to congestion information for the connecting roads.
Although there are several drawbacks to this method,
such as road segment overlaps where routes merge between stops,
this is minimal and our approach is sufficient for the current work.
Future work will investigate methods of detecting where routes merge
(physical intersections) and using these as nodes in the network.

\begin{figure}[tb]
    \centering
    \begin{subfigure}{0.7\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_1.pdf}
        \caption{Raw GTFS route shapes}
        \label{fig:network_creation_1}
    \end{subfigure} \\
    \begin{subfigure}{0.7\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_2.pdf}
        \caption{GTFS-based road network}
        \label{fig:network_creation_2}
    \end{subfigure}
    \caption{
        Constructing a route network from the stops (dots) of three separate
        routes (lines), as shown in (a).
        Routes using the same subsequence of stops are assigned the same
        road segments, so the road network shown 
        consists of nodes (stops) and edges (roads),
        as shown in (b), where the thickness of each lines
        corresponds to how many routes share that road.
    }
    \label{fig:network_creation}
\end{figure}

@@

\emph{GTFS-realtime} provides the positions of vehicles in a transit network.
The data consists of the time $t_k$ that the position was last updated,
the GPS position of the vehicle, $\bY_k$, 
and information about the trip being serviced.
In Auckland, vehicle positions are updated with a frequency of anywhere between 10~seconds and several minutes,
so there is often a lot of uncertainty about a vehicle's intermediate trajectory,
particularly when there are one or more bus stops along the way.
It is also possible for a bus to remain stationary,
for example due to heavy congestion,
so the number of possible trajectories rapidly increases with 
the time between observations.

One important consideration regarding Auckland Transport's \rt implementation is that
buses are programmed to report their location when arriving at or departing from
bus stops, as well as some major intersections.
To further complicate this,
these positions can be preemptive,
for example when approaching a queue of traffic at an intersection
and the way-point is triggered before the bus physically gets there,
so consecutive observations can show a bus travelling backwards.
To handle this, we compute the approximate distance travelled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the current state is rejected and the vehicle reverted
to its previous state before continuing.
We are available to avoid degeneration of the particle filter (Section~\ref{sec:model_perf})
at the cost of losing one observation that is most likely invalid anyway.
